Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-29 paper.tex
Closes #255 ￼

Conversion was performed on a Standard\_D32s\_v5 Azure VM instance, with 
32 CPUs and 128 GB RAM. We first converted the VCF
data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
The ICF representation required a total of 41 GiB of storage.
We then converted to VCF Zarr format using the default schema, which resulted
in a 41 GiB VCZ over 4514 chunk files. 
On inspection, this was dominated by three floating point
fields LRR (Log R Ratio), BAF (B allele frequency) and GS (GenCall Score)
stored
to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
places for GS).
On inspecting the histograms for these fields
(Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
was likely unnecessary, and we investigated applying the Quantize
and BitRound filters from \texttt{numcodecs}~\citep{numcodecs}
to reduce the precision in stored floating point data,
and therefore improve compression.
Using the Quantize filter with 5 digits of precision on these three fields reduced the
overall storage space to 31 GiB with a mean absolute error (MAE, i.e., the average
difference between the truncated and original values) of
% 3   call_LRR    quantize    1.91e-06
% 4   call_BAF    quantize    1.18e-06
% 5   call_GS quantize    1.88e-06
$1.91\times10^{-6}$ (LRR),
$1.18\times10^{-6}$ (BAF)  and
$1.88\times10^{-6}$ (GS).
Using the BitRound filter with a parameter of 5 bits reduced the overall storage for the Zarr to
17 GiB with an MAE of
% 0   call_LRR    bitround    6.93e-04
% 1   call_BAF    bitround    7.86e-04
% 2   call_GS bitround    3.38e-03
$6.93\times10^{-4}$ (LRR),
$7.86\times10^{-4}$ (BAF)  and
$3.38\times10^{-3}$ (GS).
The histograms of the truncated fields in all cases were visually indistinguishable
from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
Table~\ref{tab-ofh-data} shows a summary of the Zarr dataset using the 
BitRound filter.

@@

\begin{figure}[h]
\begin{tabular}{ccc}
\includegraphics[width=2.5cm]{ofh_example/plots/default_call_LRR.png}&
\includegraphics[width=2.5cm]{ofh_example/plots/default_call_BAF.png}&
\includegraphics[width=2.5cm]{ofh_example/plots/default_call_GS.png}
\end{tabular}
\caption{Distribution of floating point fields in the Our Future Health data.
\label{fig-ofh-field-distributions}}
\end{figure}

>>>>

Conversion was performed on a Standard\_D32s\_v5 Azure VM instance, with 
32 CPUs and 128 GB RAM. We first converted the VCF
data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
The ICF representation required a total of 41 GiB of storage.
We then converted to VCF Zarr format using the default schema, which resulted
in a 41 GiB VCZ over 4514 chunk files. 
On inspection, this was dominated by three floating point
fields LRR (Log R Ratio), BAF (B allele frequency) and GS (GenCall Score)
stored
to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
places for GS).
On inspecting the histograms for these fields
(Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
was likely unnecessary, and we investigated applying the Quantize
and BitRound filters from \texttt{numcodecs}~\citep{numcodecs}
to reduce the precision in stored floating point data,
and therefore improve compression.
Using the Quantize filter with 5 digits of precision on these three fields reduced the
overall storage space to 31 GiB with a mean absolute error (MAE, i.e., the average
difference between the truncated and original values) of
% 3   call_LRR    quantize    1.91e-06
% 4   call_BAF    quantize    1.18e-06
% 5   call_GS quantize    1.88e-06
$1.91\times10^{-6}$ (LRR),
$1.18\times10^{-6}$ (BAF)  and
$1.88\times10^{-6}$ (GS).
Using the BitRound filter with a parameter of 5 bits reduced the overall storage for the Zarr to
17 GiB with an MAE of
% 0   call_LRR    bitround    6.93e-04
% 1   call_BAF    bitround    7.86e-04
% 2   call_GS bitround    3.38e-03
$6.93\times10^{-4}$ (LRR),
$7.86\times10^{-4}$ (BAF)  and
$3.38\times10^{-3}$ (GS).
The histograms of the truncated fields in all cases were very similar to the
to the original (Fig~\ref{fig-ofh-field-distributions}).
Table~\ref{tab-ofh-data} shows a summary of the Zarr dataset using the 
BitRound filter.

@@

\begin{figure}[h]
\includegraphics{figures/ofh-histograms}
\caption{Distribution of values in the GS, LRR and BAF fields in the Our Future
Health data with no truncation (Default), and truncation using the BitRound
and Quantize filters.
\label{fig-ofh-field-distributions}}
\end{figure}
